VOI-aware Monte-Carlo Sampling in Trees

### 2011-12-10 paper/vago/paper.tex
2 in blinkered is back, that was a temporary blackout, sorry

The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be})
can be bounded from above using concentration inequalities. In
particular, Lemma~\ref{lemma:hoeffding-prob-bounds} (proved in 
Appendix~\ref{app:hoeffding-prob-bounds-proof}) is based on the
Hoeffding inequality (\ref{eqn:conc-hoeffding}):
\begin{lmm} The probabilities in equations (\ref{eqn:thm-vopi},
\ref{eqn:thm-be}) are bounded from above as
\begin{eqnarray}
\Pr(\IE[X_i] \le \overline X_\beta|i=\alpha)&
   \le & \exp(-2 (\overline X_i - \overline X_\beta)^2 n_i)\nonumber\\
\Pr(\IE[X_i] \ge \overline X_\alpha|i\ne\alpha)&
   \le & \exp(-2 (\overline X_\alpha - \overline X_i)^2 n_i)
\label{eqn:probound-perf-hoeffding}
\end{eqnarray}
\vspace{-1.5\baselineskip}
\begin{eqnarray}
\Pr(\overline X_i' \le \overline X_\beta|i=\alpha) &
   \le & \exp\left(- 2\left(\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}}(\overline X_i - \overline X_\beta)\right)^2 n_i \right)
   \le \exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i \right) \nonumber\\
\Pr(\overline X_i' \ge \overline X_\alpha|i\ne\alpha) &
   \le & \exp\left(- 2\left(\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}} (\overline X_\alpha - \overline X_i)\right)^2 n_i \right)
    \le \exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i \right)
\label{eqn:probound-blnk-hoeffding}
\end{eqnarray}
\label{lemma:hoeffding-prob-bounds}
\end{lmm}

An upper bound on the intrinsic value of perfect information is obtained by substituting
(\ref{eqn:probound-perf-hoeffding}) into (\ref{eqn:thm-vopi}):
\begin{equation}
  \Lambda_i^p \le \hat\Lambda_i^p=\left\{
  \begin{array}{l l}
    \overline X_\beta\exp\left(- 2(\overline X_i - \overline X_\beta)^2 n_i \right) & \mbox{{\rm if} $i=\alpha$} \\
    (1-\overline  X_\alpha)\exp\left(- 2(\overline X_\alpha - \overline X_i)^2 n_i \right)  &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-perf-hoeffding}
\end{equation}
An upper bound on the blinkered VOI estimate with ordering independent of $N$ is obtained
by substituting (\ref{eqn:probound-blnk-hoeffding}) into (\ref{eqn:thm-be}).
\begin{equation}
  \Lambda_i^b \le \hat\Lambda_i^b=\left\{
  \begin{array}{l l}
    N \frac {\overline X_\beta} {n_i}\exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
    N \frac {(1-\overline  X_\alpha)} {n_i} \exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i\right)
      &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-blnk-hoeffding}
\end{equation}
The bound is generally tighter when $N$ is small relative to $n_i$,
and the estimate $\hat\Lambda_i^b$ can be viewed as a relaxation of the myopic VOI
estimate to longer sequences of actions, such
that the VOI estimate is positive for some of the actions.

@@

The experiments compare the UCT algorithm \cite{Kocsis.uct} with modified
versions of UCT in which the samples at the first step are selected according
to their VOI estimates. Three blinkered VOI estimates were used for the comparison:
\begin{description}
\item[VCT:] based on Hoeffding inequality (\ref{eqn:conc-hoeffding});
\item[ECT:] based on Hoeffding inequality with midpoint (see Appendix~\ref{app:better-hoeffding-bound});
\item[BCT:] based on empirical Bernstein inequality (\ref{eqn:conc-empbernstein}).
\end{description}
For simplicity, the bounds on probability for unlimited number of samples were used in the estimate formulas. The estimates where computed as follows:
\begin{eqnarray}\
  \Lambda_i^{VCT}& =&\left\{
  \begin{array}{l l}
    \frac {\overline X_\beta} {n_i}\exp\left(- 2(\overline X_i - \overline X_\beta)^2 n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
    \frac {(1-\overline  X_\alpha)} {n_i}\exp\left(- 2(\overline X_\alpha - \overline X_i)^2 n_i\right)
      &  \mbox{\rm otherwise}
  \end{array}  \right.\nonumber\\
  \Lambda_i^{ECT}& =&\min\left(\Lambda_i^{VCT}, \min_y\left\{
  \begin{array}{l l}
    \frac {\overline X_\beta-y} {n_i}\exp\left(-2 (\overline X_i-\overline X_\beta)^2n_i\right)
    +\frac {\overline X_\beta} {n_i}\exp\left(-2 (\overline X_i-y)^2n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
\frac {y-\overline X_\alpha} {n_i}\exp\left(-2(\overline X_\alpha-\overline X_i)^2n_i\right)
    +\frac {1-\overline X_\alpha} {n_i}\exp\left(-2 (y-\overline X_i)^2n_i\right)
      &  \mbox{\rm otherwise}
  \end{array} \right.\right) \nonumber\\
  \Lambda_i^{BCT}& =&\min\left(\Lambda_i^{VCT},\left\{
  \begin{array}{l l}
    2\frac {\overline X_\beta} {n_i}\exp\left(- \frac {(\overline X_i - \overline X_\beta)^2 n_i} {\frac {14} 3 \frac {n_i} {n_i-1}(\overline X_i - \overline X_\beta) + 2 \overline X_i(1-\overline X_i)} \right)
      & \mbox{{\rm if} $i=\alpha$} \\
    2\frac {(1-\overline  X_\alpha)} {n_i}\exp\left(- \frac {(\overline X_\alpha - \overline X_i)^2 n_i} {\frac {14} 3 \frac {n_i} {n_i-1}(\overline X_\alpha - \overline X_i) + 2 \overline X_i(1-\overline X_i)} \right)
      &  \mbox{\rm otherwise}
  \end{array} \right.\right)
\label{eqn:emp-estimates}
\end{eqnarray}

@@

For any $\delta$, the probability that $\overline X_i'$ is greater
than $\overline X_\alpha$ is less than the probability that $\IE[X_i]\ge\overline
X_i+\delta$ \emph{or} $\overline X_i'\le \IE[X_i]+\overline X_\alpha - \overline
  X_i - \delta +(\overline X_\alpha - \overline X_i)\frac {n_i} N$, thus
less than the maximum of probabilities of the two events:
\begin{equation}
\Pr(\overline X_i'\ge \overline X_\alpha)\le
  \max\left(\Pr(\IE[X_i]-\overline X_i \ge \delta),\,
  \Pr\left(\overline X_i^+ - \IE[X_i] \ge \overline X_\alpha - \overline
  X_i - \delta +(\overline X_\alpha - \overline X_i)\frac {n_i} N\right)\right)
\end{equation}
Bounding the probabilities on the right side using the Hoeffding
inequality, obtain:
\begin{equation}
\Pr(\overline X_i'\ge \overline X_\alpha)\le 
\max\left(\exp(-2\delta^2n_i),\,
          \exp\left(-2\left((\overline X_\alpha 
          - \overline X_i)\left(1+\frac {n_i} N\right)
          - \delta\right)^2N\right)\right)
\label{eqn:app-hoeffding-le-maxexp}
\end{equation}
The maximum is equal to either of the arguments when the arguments are equal. Let
\begin{equation}
\exp(-\delta^2n) = \exp(-2\left((\overline X_\alpha - \overline X_i)(1+\frac {n_i} N) - \delta\right)^2N)\label{eqn:app-hoeffding-eq-exp}
\end{equation}
Solve (\ref{eqn:app-hoeffding-eq-exp}) for $\delta$: $\delta=\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}} (\overline X_\alpha
- \overline X_i) \ge (\sqrt 2 - 1)(\overline X_\alpha-\overline X_i)$. Substitute $\delta$ into 
(\ref{eqn:app-hoeffding-le-maxexp}) and obtain
\begin{eqnarray}
\Pr(\overline X_i'\ge \overline X_\alpha) 
& \le & \exp\left(-2\left( \frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}}
                          (\overline X_\alpha - \overline X_i)\right)^2 n_i\right)\nonumber \\
& \le & \exp(-2(\sqrt 2 - 1)^2(\overline X_\alpha - \overline X_i)^2n_i)
 \le \exp(-1.37(\overline X_\alpha - \overline X_i)^2n_i)
\end{eqnarray}
Derivation for the case $i=\alpha$ is similar.
\end{proof}

>>>>

The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be})
can be bounded from above using concentration inequalities. In
particular, Lemma~\ref{lemma:hoeffding-prob-bounds} (proved in 
Appendix~\ref{app:hoeffding-prob-bounds-proof}) is based on the
Hoeffding inequality (\ref{eqn:conc-hoeffding}):
\begin{lmm} The probabilities in equations (\ref{eqn:thm-vopi},
\ref{eqn:thm-be}) are bounded from above as
\begin{eqnarray}
\Pr(\IE[X_i] \le \overline X_\beta|i=\alpha)&
   \le & \exp(-2 (\overline X_i - \overline X_\beta)^2 n_i)\nonumber\\
\Pr(\IE[X_i] \ge \overline X_\alpha|i\ne\alpha)&
   \le & \exp(-2 (\overline X_\alpha - \overline X_i)^2 n_i)
\label{eqn:probound-perf-hoeffding}
\end{eqnarray}
\vspace{-1.5\baselineskip}
\begin{eqnarray}
\Pr(\overline X_i' \le \overline X_\beta|i=\alpha) &
   \le & 2\exp\left(- 2\left(\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}}(\overline X_i - \overline X_\beta)\right)^2 n_i \right)
   \le 2\exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i \right) \nonumber\\
\Pr(\overline X_i' \ge \overline X_\alpha|i\ne\alpha) &
   \le & 2\exp\left(- 2\left(\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}} (\overline X_\alpha - \overline X_i)\right)^2 n_i \right)
    \le 2\exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i \right)
\label{eqn:probound-blnk-hoeffding}
\end{eqnarray}
\label{lemma:hoeffding-prob-bounds}
\end{lmm}

An upper bound on the intrinsic value of perfect information is obtained by substituting
(\ref{eqn:probound-perf-hoeffding}) into (\ref{eqn:thm-vopi}):
\begin{equation}
  \Lambda_i^p \le \hat\Lambda_i^p=\left\{
  \begin{array}{l l}
    \overline X_\beta\cdot \exp\left(- 2(\overline X_i - \overline X_\beta)^2 n_i \right) & \mbox{{\rm if} $i=\alpha$} \\
    (1-\overline  X_\alpha)\cdot \exp\left(- 2(\overline X_\alpha - \overline X_i)^2 n_i \right)  &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-perf-hoeffding}
\end{equation}
An upper bound on the blinkered VOI estimate with ordering independent of $N$ is obtained
by substituting (\ref{eqn:probound-blnk-hoeffding}) into (\ref{eqn:thm-be}).
\begin{equation}
  \Lambda_i^b \le \hat\Lambda_i^b=\left\{
  \begin{array}{l l}
    N\frac {\overline X_\beta} {n_i}\cdot 2\exp\left(- 1.37(\overline X_i - \overline X_\beta)^2 n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
    N\frac {(1-\overline  X_\alpha)} {n_i}\cdot 2\exp\left(- 1.37(\overline X_\alpha - \overline X_i)^2 n_i\right)
      &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-blnk-hoeffding}
\end{equation}
The bound is generally tighter when $N$ is small relative to $n_i$,
and the estimate $\hat\Lambda_i^b$ can be viewed as a relaxation of the myopic VOI
estimate to longer sequences of actions, such
that the VOI estimate is positive for some of the actions.

@@

The experiments compare the UCT algorithm \cite{Kocsis.uct} with modified
versions of UCT in which the samples at the first step are selected according
to their VOI estimates. Three blinkered VOI estimates were used for the comparison:
\begin{description}
\item[VCT:] based on Hoeffding inequality (\ref{eqn:conc-hoeffding});
\item[ECT:] based on Hoeffding inequality with midpoint (see Appendix~\ref{app:better-hoeffding-bound});
\item[BCT:] based on empirical Bernstein inequality (\ref{eqn:conc-empbernstein}).
\end{description}
For simplicity, the bounds on probability for unlimited number of samples were used in the estimate formulas. The estimates where computed as follows:
\begin{eqnarray}\
  \Lambda_i^{VCT}& =&\left\{
  \begin{array}{l l}
    \frac {\overline X_\beta} {n_i} \cdot \exp\left(- 2(\overline X_i - \overline X_\beta)^2 n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
    \frac {(1-\overline  X_\alpha)} {n_i}\cdot \exp\left(- 2(\overline X_\alpha - \overline X_i)^2 n_i\right)
      &  \mbox{\rm otherwise}
  \end{array}  \right.\nonumber\\
  \Lambda_i^{ECT}& =&\min\left(\Lambda_i^{VCT}, \min_y\left\{
  \begin{array}{l l}
    \frac {\overline X_\beta-y} {n_i}\cdot \exp\left(-2 (\overline X_i-\overline X_\beta)^2n_i\right)
    +\frac {\overline X_\beta} {n_i}\cdot \exp\left(-2 (\overline X_i-y)^2n_i\right)
      & \mbox{{\rm if} $i=\alpha$} \\
\frac {y-\overline X_\alpha} {n_i}\cdot \exp\left(-2(\overline X_\alpha-\overline X_i)^2n_i\right)
    +\frac {1-\overline X_\alpha} {n_i}\cdot \exp\left(-2 (y-\overline X_i)^2n_i\right)
      &  \mbox{\rm otherwise}
  \end{array} \right.\right) \nonumber\\
  \Lambda_i^{BCT}& =&\min\left(\Lambda_i^{VCT},\left\{
  \begin{array}{l l}
    \frac {\overline X_\beta} {n_i}\cdot 2\exp\left(- \frac {(\overline X_i - \overline X_\beta)^2 n_i} {\frac {14} 3 \frac {n_i} {n_i-1}(\overline X_i - \overline X_\beta) + 2 \overline X_i(1-\overline X_i)} \right)
      & \mbox{{\rm if} $i=\alpha$} \\
    \frac {(1-\overline  X_\alpha)} {n_i}\cdot 2\exp\left(- \frac {(\overline X_\alpha - \overline X_i)^2 n_i} {\frac {14} 3 \frac {n_i} {n_i-1}(\overline X_\alpha - \overline X_i) + 2 \overline X_i(1-\overline X_i)} \right)
      &  \mbox{\rm otherwise}
  \end{array} \right.\right)
\label{eqn:emp-estimates}
\end{eqnarray}

@@

For any $\delta$, the probability that $\overline X_i'$ is greater
than $\overline X_\alpha$ is less than the probability that
$\IE[X_i]\ge\overline X_i+\delta$
\emph{or} $\overline X_i^+\ge \IE[X_i]+\overline X_\alpha
- \overline X_i - \delta +(\overline X_\alpha - \overline X_i)\frac {n_i} N$,
thus, by the union bound, less than the sum of the probabilities:
\begin{equation}
\Pr(\overline X_i'\ge \overline X_\alpha)\le
  \Pr(\IE[X_i]-\overline X_i \ge \delta)+
  \Pr\left(\overline X_i^+ - \IE[X_i] \ge \overline X_\alpha
           - \overline X_i - \delta +(\overline X_\alpha - \overline X_i)\frac {n_i} N\right)
\end{equation}
Bounding the probabilities on the right-hand side using the Hoeffding
inequality, obtain:
\begin{equation}
\Pr(\overline X_i'\ge \overline X_\alpha)\le 
    \exp(-2\delta^2n_i)+
         \exp\left(-2\left((\overline X_\alpha 
         - \overline X_i)\left(1+\frac {n_i} N\right)
         - \delta\right)^2N\right)
\label{eqn:app-hoeffding-le-maxexp}
\end{equation}
Find $\delta$ for which the two terms on the right-hand side of
(\ref{eqn:app-hoeffding-le-maxexp}) are equal:
\begin{equation}
\exp(-\delta^2n) = \exp(-2\left((\overline X_\alpha - \overline X_i)(1+\frac {n_i} N) - \delta\right)^2N)\label{eqn:app-hoeffding-eq-exp}
\end{equation}
Solve (\ref{eqn:app-hoeffding-eq-exp}) for $\delta$: $\delta=\frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}} (\overline X_\alpha
- \overline X_i) \ge (\sqrt 2 - 1)(\overline X_\alpha-\overline X_i)$. Substitute $\delta$ into 
(\ref{eqn:app-hoeffding-le-maxexp}) and obtain
\begin{eqnarray}
\Pr(\overline X_i'\ge \overline X_\alpha) 
& \le & 2\exp\left(-2\left( \frac {1+\frac {n_i} N} {1+\sqrt {\frac {n_i} N}}
                          (\overline X_\alpha - \overline X_i)\right)^2 n_i\right)\nonumber \\
& \le & 2\exp(-2(\sqrt 2 - 1)^2(\overline X_\alpha - \overline X_i)^2n_i)
 \le 2\exp(-1.37(\overline X_\alpha - \overline X_i)^2n_i)
\end{eqnarray}
Derivation for the case $i=\alpha$ is similar.
\end{proof}
